---
title: "QCB 408 / 508 -- Notes on Week X"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

\providecommand{\E}{\operatorname{E}}
\providecommand{\V}{\operatorname{Var}}
\providecommand{\Cov}{\operatorname{Cov}}
\providecommand{\se}{\operatorname{se}}
\providecommand{\logit}{\operatorname{logit}}
\providecommand{\iid}{\; \stackrel{\text{iid}}{\sim}\;}
\providecommand{\asim}{\; \stackrel{.}{\sim}\;}
\providecommand{\xs}{x_1, x_2, \ldots, x_n}
\providecommand{\Xs}{X_1, X_2, \ldots, X_n}
\providecommand{\bB}{\boldsymbol{B}}
\providecommand{\bb}{\boldsymbol{\beta}}
\providecommand{\bx}{\boldsymbol{x}}
\providecommand{\bX}{\boldsymbol{X}}
\providecommand{\by}{\boldsymbol{y}}
\providecommand{\bY}{\boldsymbol{Y}}
\providecommand{\bz}{\boldsymbol{z}}
\providecommand{\bZ}{\boldsymbol{Z}}
\providecommand{\be}{\boldsymbol{e}}
\providecommand{\bE}{\boldsymbol{E}}
\providecommand{\bs}{\boldsymbol{s}}
\providecommand{\bS}{\boldsymbol{S}}
\providecommand{\bP}{\boldsymbol{P}}
\providecommand{\bI}{\boldsymbol{I}}
\providecommand{\bD}{\boldsymbol{D}}
\providecommand{\bd}{\boldsymbol{d}}
\providecommand{\bW}{\boldsymbol{W}}
\providecommand{\bw}{\boldsymbol{w}}
\providecommand{\bM}{\boldsymbol{M}}
\providecommand{\bPhi}{\boldsymbol{\Phi}}
\providecommand{\bphi}{\boldsymbol{\phi}}
\providecommand{\bN}{\boldsymbol{N}}
\providecommand{\bR}{\boldsymbol{R}}
\providecommand{\bu}{\boldsymbol{u}}
\providecommand{\bU}{\boldsymbol{U}}
\providecommand{\bv}{\boldsymbol{v}}
\providecommand{\bV}{\boldsymbol{V}}
\providecommand{\bO}{\boldsymbol{0}}
\providecommand{\bOmega}{\boldsymbol{\Omega}}
\providecommand{\bLambda}{\boldsymbol{\Lambda}}
\providecommand{\bSig}{\boldsymbol{\Sigma}}
\providecommand{\bSigma}{\boldsymbol{\Sigma}}
\providecommand{\bt}{\boldsymbol{\theta}}
\providecommand{\bT}{\boldsymbol{\Theta}}
\providecommand{\bpi}{\boldsymbol{\pi}}
\providecommand{\argmax}{\text{argmax}}
\providecommand{\KL}{\text{KL}}
\providecommand{\fdr}{\text{FDR}}
\providecommand{\pfdr}{{\rm pFDR}}
\providecommand{\mfdr}{{\rm mFDR}}
\providecommand{\bh}{\hat}
\providecommand{\dd}{\lambda}
\providecommand{\q}{\operatorname{q}}

```{r, message=FALSE, echo=FALSE, cache=FALSE}
# this makes the R output formatted the same as Foundations of Applied Statistics
source("https://raw.githubusercontent.com/jdstorey/fas/master/customization/knitr_options.R")

# load libraries used in this chunk
library(gapminder) 
```

If your name is Robert Fripp and you're scribing notes for Week 3, for example, then rename this file `week03_scribed_fripp.Rmd`. Any additional files, such as data files should be included in a directory called `data/`, where this directory is in the same location as `week03_scribed_fripp.Rmd`. The data would then be accessed below by `./data/dataset.csv`, for example. Do not deviate from this convention. Also, do not use a data file name that has already been used (see [here](https://github.com/jdstorey/fsg/tree/master/docs/scribed_notes/data/) for those already in use). By default the working directory when compiling a `.Rmd` file is the location of the `.Rmd` file. 

## Summary

A summary of the week's topics discussed below (like an abstract). Maybe you want to include a list of the topics as well:

- Probability
- Random Variables
- Hardy-Weinberg equilibrium 

## First topic

There are $\LaTeX$ macros above to make each person's notes similar in format. Examples:
$$
\E[X] = \sum_{k} k \Pr(X=k),
$$
$$
\fdr = \E \left[ \frac{V}{R \vee 1} 1(R > 0) \right].
$$

Of course, equations can be inline as well, $n \rightarrow \infty$.

### First subsection

```{r}
before1980 <- gapminder %>% filter(year < 1980) %>% 
  select(lifeExp) %>% unlist()
after1980 <- gapminder %>% filter(year > 1980) %>% 
  select(lifeExp) %>% unlist()
qqplot(before1980, after1980); abline(0,1)
```

### Second subsection

## Second topic

### First subsection

### Second subsection

## Etc.

## Session Information

```{r}
sessionInfo()
```

  
  
  

